---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE
)
```
# mermaidr

<!-- badges: start -->
<!-- badges: end -->

The goal of `mermaidr` is to access [MERMAID Collect](https://collect.datamermaid.org/) data directly from R. The package is presently set to access the [dev MERMAID Collect](https://dev-collect.datamermaid.org/) only.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mermaidr/mermaidr@package")
```

## Authentication

`mermaidr` will help you interact with MERMAID Collect as an authenticated user, as soon as you need. This is only required for accessing project specific data. To access a list of projects, sites, etc, you do not need to be authenticated.

If you would like to authenticate yourself immediately, use `mermaid_auth()`. This will open your browser to the MERMAID Collect login. Once you log in, you can go back to R and will be authenticated.

The login credentials expire every 24 hours. Once your credentials are expired, `mermaidr` will again help you automatically authenticate when needed.

## Usage

To access the unauthenticated API endpoints, use `get_mermaid_endpoints()`. The results will return as a `tibble.` The following endpoints are available: "benthicattributes", "fishattributes", "fishfamilies", "fishgenera", "fishspecies", "managements", "projects", "sites".

For example,

```{r}
library(mermaidr)

get_mermaid_endpoint("projects")
```

```{r}
get_mermaid_endpoint("managements")
```

### Accessing project data

You will be able to access data from a specific project, provided that you have access to it in the Collect app. To access data for a project, you can either use a project from `get_mermaid_endpoint("projects")` (as above), a `project_id` directly, or a project from `search_projects()`. 

For example:

```{r}
library(dplyr)

mermaidr_project <- search_projects(name = "mermaidr testing", exact_name = TRUE)

mermaidr_project
```

returns a single project with the exact name "mermaidr testing".

You can use this to access an endpoint for the project, using `get_mermaid_project_endpoint()`. The following project endpoints are available: "beltfishtransectmethods", "beltfishes", "benthiclittransectmethods", "benthicpittransectmethods", "benthicpits", "collectrecords", "habitatcomplexities", "obsbenthiclits", "obsbenthicpits", "obshabitatcomplexities", "obstransectbeltfishs", "managements", "observers", "project_profiles", "sampleevents", "sites".

For example, to see the sites in this project:

```{r}
get_mermaid_project_endpoint(mermaidr_project, "sites")
```

You can also use the `project_id` directly to access data from a project, without having to search for it first. This may be handy since the `project_id` is directly available from the URL when using the collect app.

```{r}
get_mermaid_project_endpoint("e477b009-cfd9-4d71-9b8c-d1684f38b954", "managements")
```

If you want to access data from the same project multiple times within a session, it may be useful to set the default project, rather than having to supply it every time. You can do this using `set_default_project()`. Then, you can just supply the endpoint, and the default project is used.

```{r}
set_default_project(mermaidr_project)
get_mermaid_project_endpoint(endpoint = "beltfishes")
```

