---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  cache = TRUE
)
```
# mermaidr

<!-- badges: start -->
[![CircleCI build status](https://circleci.com/gh/data-mermaid/mermaidr.svg?style=svg)](https://circleci.com/gh/data-mermaid/mermaidr)
<!-- badges: end -->

The goal of `mermaidr` is to access [MERMAID Collect](https://collect.datamermaid.org/) data directly from R. It works alongside the [`mermaidreporting` package](https://github.com/data-mermaid/mermaidreporting), which helps to clean, summarize, and visualize MERMAID data.

If you run into any problems working with this package, please open an [issue](https://github.com/data-mermaid/mermaidr/issues).

## Installation

You can install mermaidr from GitHub with:

``` r
# install.packages("remotes")
remotes::install_github("data-mermaid/mermaidr", upgrade = "never")
```

Next, load the package:

```{r}
library(mermaidr)
```

If you would like to access the development version of MERMAID instead, you can install the `dev` branch of this package:

``` r
# install.packages("remotes")
remotes::install_github("data-mermaid/mermaidr", ref = "dev", upgrade = "never")
```

When using the development version, you can only access data from the [development version of MERMAID Collect](https://dev-collect.datamermaid.org/). There may also be differences in the MERMAID API (which can affect things like the columns returned) and functions in `mermaidr` that are in-progress and not yet available from the "production" version of the package.

## Authentication

`mermaidr` will help you interact with MERMAID Collect as an authenticated user, as soon as you need. This is only required for accessing project specific data. To access a list of all projects, sites, etc, you do not need to be authenticated.

If you would like to authenticate yourself immediately, use `mermaid_auth()`. This will open your browser to the MERMAID Collect login. Once you log in, you can go back to R and will be authenticated.

The login credentials expire every 24 hours. Once your credentials are expired, `mermaidr` will again help you automatically authenticate when needed.

Note that authentication is only possible locally on your desktop, using a program like RStudio. This should cover most people and cases, however it's not currently possible to authenticate on something like RStudio Cloud or on another server. If you are using a server, please follow the directions [here](https://support.rstudio.com/hc/en-us/articles/217952868-Generating-OAuth-tokens-from-a-server) to authenticate on the desktop then copy to the server.

## Usage

All functions in `mermaidr` are of the form `mermaid_*()`, to make functions easier to find and use when loaded with other packages!

## Accessing project data

To access data related to your MERMAID projects, first obtain a list of your projects with `mermaid_get_my_projects()`. 

At this point, you will have to authenticate to the Collect app. R will help you do this automatically by opening a browser window for you to log in to Collect, either via Google sign-in or username and password - however you normally do!

Once you've logged in, come back to R. Your login credentials will be stored for a day, until they expire, and you will need to login again. The package handles the expiration for you, so just log in again when prompted.

```{r}
library(mermaidr)
my_projects <- mermaid_get_my_projects()

my_projects
```

This function returns information on your projects, including project countries, the number of sites, tags, data policies, and more.

To filter for specific projects, you can use the `filter` function from `dplyr`:

```{r, message = FALSE}
library(dplyr)

indonesia_projects <- my_projects %>%
  filter(countries == "Indonesia")

indonesia_projects
```

Alternatively, you can search your projects using `mermaid_search_my_projects()`, narrowing projects down by name, countries, or tags:

```{r}
mermaid_search_my_projects(countries = "Indonesia")
```

Then, you can start to access data about your projects, like project sites via `mermaid_get_project_sites()`:

```{r}
indonesia_projects %>%
  mermaid_get_project_sites()
```

Or the managements for your projects via `mermaid_get_project_managements()`:

```{r}
indonesia_projects %>%
  mermaid_get_project_managements()
```

### Method data

You can also  access data on your projects' Fish Belt, Benthic LIT, Benthic PIT, Bleaching, and Habitat Complexity methods. The details are in the following sections.

#### Fish Belt data

To access Fish Belt data for a project, use `mermaid_get_project_data()` with `method = "fishbelt"`. 
You can access individual observations (i.e., a record of each observation) by setting `data = "observations"`:

```{r}
xpdc <- my_projects %>%
  filter(name == "XPDC Kei Kecil 2018")

xpdc %>%
  mermaid_get_project_data(method = "fishbelt", data = "observations")
```

You can access sample units data, which are observations aggregated to the sample units level. Fish belt sample units contain total biomass in kg/ha per sample unit, by trophic group:

```{r}
xpdc %>%
  mermaid_get_project_data("fishbelt", "sampleunits")
```

And finally, sample events data, which are aggregated further, to the sample event level. Fish belt sample events contain *mean* total biomass in kg/ha per sample event and by trophic group:

```{r}
xpdc_sample_events <- xpdc %>%
  mermaid_get_project_data("fishbelt", "sampleevents")

xpdc_sample_events
```

##### A note on saving data

Both the sample units and sample events data contain a "data-frame column" (`biomass_kgha_by_trophic_group` and `biomass_kgha_by_trophic_group_avg`, respectively). In order to save this data in a file like a CSV or XLSX, you will need to expand this column first.

This can be done with a function from the `mermaidreporting` package, `mermaid_clean_columns()`:

```{r}
library(mermaidreporting)

xpdc_sample_events_clean <- xpdc_sample_events %>%
  mermaid_clean_columns()

xpdc_sample_events_clean
```

Then, you can save the data:

```{r, eval = FALSE}
library(readr)

write_csv(xpdc_sample_events_clean, "xpdc_sample_events_clean.csv")
```

#### Benthic LIT data

To access Benthic LIT data, use `mermaid_get_project_data()` with `method = "benthiclit"`. 

```{r}
mozambique <- my_projects %>%
  filter(name == "WCS Mozambique Coral Reef Monitoring")

mozambique %>%
  mermaid_get_project_data(method = "benthiclit", data = "observations")
```

You can access sample units and sample events the same way. 

For Benthic LIT, sample units contain percent cover per sample unit, by benthic category. Sample *events* contain *mean* percent cover per sample event, by benthic category. 

```{r}
mozambique %>%
  mermaid_get_project_data(method = "benthiclit", data = "sampleunits")
```

#### Benthic PIT data

To access Benthic LIT data, change `method` to "benthicpit":

```{r}
xpdc %>%
  mermaid_get_project_data(method = "benthicpit", data = "observations")
```

You can access sample units and sample events the same way, and the data format is the same as Benthic LIT.

You can return both sample units and sample events by setting the `data` argument. This will return a list of two data frames: one containing sample units, and the other sample events.

```{r}
xpdc_sample_units_events <- xpdc %>%
  mermaid_get_project_data(method = "benthicpit", data = c("sampleunits", "sampleevents"))

names(xpdc_sample_units_events)
xpdc_sample_units_events[["sampleunits"]]
```

#### Bleaching

To access Bleaching data, set `method` to "bleaching". There are two types of observations data for the Bleaching method: Colonies Bleached and Percent Cover. These are both returned when pulling observations data, in a list:

```{r}
bleaching_obs <- mozambique %>%
  mermaid_get_project_data("bleaching", "observations")

names(bleaching_obs)

bleaching_obs[["colonies_bleached"]]
```

The sample units and sample events data contain summaries of both Colonies Bleached and Percent Cover:

```{r}
mozambique %>%
  mermaid_get_project_data("bleaching", "sampleevents")
```

#### Habitat Complexity

Finally, to access Habitat Complexity data, set `method` to "habitatcomplexity". As with all other methods, you can access observations, sample units, and sample events:

```{r}
xpdc %>%
  mermaid_get_project_data("habitatcomplexity", "sampleevents")
```

#### Multiple methods data

To pull data for both fish belt and benthic PIT methods, you can set `method` to include both.

```{r}
xpdc_sample_events <- xpdc %>%
  mermaid_get_project_data(method = c("fishbelt", "benthicpit"), data = "sampleevents")
```

The result is a list of data frames, containing sample events for both fish belt and benthic PIT methods:

```{r}
names(xpdc_sample_events)

xpdc_sample_events[["benthicpit"]]
```

Alternatively, you can set `method` to "all" to pull for all methods! Similarly, you can set `data` to "all" to pull all types of data:

```{r}
all_project_data <- xpdc %>%
  mermaid_get_project_data(method = "all", data = "all", limit = 1)

names(all_project_data)

names(all_project_data[["benthicpit"]])
```

#### Multiple projects

Pulling data for multiple projects is the exact same, except there will be an additional "project" column at the beginning to distinguish which projects the data comes from. Recall that `my_projects` contains six projects:

```{r}
my_projects
```

```{r}
my_projects %>%
  mermaid_get_project_data("fishbelt", "sampleevents", limit = 1)
```

Note the `limit` argument here, which just limits the data pulled to one record (per project, method, and data combination). This is useful if you want to get a preview of what your data will look like without having to pull it all in.

### Accessing non-project data

You may also want to access data that is not related to projects. To access this data, you do not need to authenticate R with MERMAID.

For example, you can pull reference data (the names and information of the fish and benthic attributes you can choose in MERMAID), using `mermaid_get_reference()`:

```{r}
mermaid_get_reference(reference = "fishfamilies")
```

Using this function, you can access the fish family, fish genera, fish species, and benthic attributes references by changing the `reference` argument.

You can also get a list of *all* projects (not just your own):

```{r}
mermaid_get_projects()
```

As well as all sites:

```{r}
mermaid_get_sites()
```

And all managements:

```{r}
mermaid_get_managements()
```

### Other data

There is additional data available from the MERMAID API, both related to specific projects and not. If you think you'll need to use these, please see the help for them by typing `?mermaid_get_endpoint` or `?mermaid_get_project_endpoint`.
