---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE
)
```
# mermaidr

<!-- badges: start -->
<!-- badges: end -->

The goal of `mermaidr` is to access [MERMAID Collect](https://collect.datamermaid.org/) data directly from R. The package is presently set to access the [dev MERMAID Collect](https://dev-collect.datamermaid.org/) only.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mermaidr/mermaidr@package")
```

## Authentication

`mermaidr` will help you interact with MERMAID Collect as an authenticated user, as soon as you need. This is only required for accessing project specific data. To access a list of projects, sites, etc, you do not need to be authenticated.

If you would like to authenticate yourself immediately, use `mermaid_auth()`. This will open your browser to the MERMAID Collect login. Once you log in, you can go back to R and will be authenticated.

The login credentials expire every 24 hours. Once your credentials are expired, `mermaidr` will again help you automatically authenticate when needed.

## Usage

To access the unauthenticated API endpoints, use `get_mermaid_endpoints()`. The results will return as a `tibble.` The following endpoints are available: "benthicattributes", "fishattributes", "fishfamilies", "fishgenera", "fishspecies", "managements", "projects", "sites".

For example,

```{r}
library(mermaidr)

get_mermaid_endpoint("projects")
```

By default, the function returns 50 results - to get more (or less!), use the `limit` argument:

```{r}
get_mermaid_endpoint("managements", limit = 5)
```

### Accessing project data

You will be able to access data from a specific project, provided that you have access to it in the Collect app. To access data for a project, you can either use a project from `get_mermaid_endpoint("projects")` (as above), a `project_id` directly, or a project from `search_projects()`. 

For example:

```{r}
mermaidr_project <- search_projects(name = "Sharla test")

mermaidr_project
```

returns a single project with the exact name "Sharla test".

You can use this to access an endpoint for the project, using `get_mermaid_project_endpoint()`. The following project endpoints are available: "beltfishtransectmethods", "beltfishes", "benthiclittransectmethods", "benthicpittransectmethods", "benthicpits", "collectrecords", "habitatcomplexities", "obsbenthiclits", "obsbenthicpits", "obshabitatcomplexities", "obstransectbeltfishs", "managements", "observers", "project_profiles", "sampleevents", "sites".

At this point, you will have to authenticate to the Collect app. R will help you do this automatically by opening a browser window for you to log in to Collect, either via Google sign-in or username and password - however you normally do!

Once you've logged in, come back to R. Your login credentials will be stored for a day, until they expire, and you will need to login again. The package handles the expiration for you, so just log in again when prompted.

For example, to see the sites in this project:

```{r}
get_mermaid_project_endpoint("sites", mermaidr_project)
```

You can also use the `project_id` directly to access data from a project, without having to search for it first. This may be handy since the `project_id` is directly available from the URL when using the collect app.

```{r}
get_mermaid_project_endpoint("managements", "2c0c9857-b11c-4b82-b7ef-e9b383d1233c")
```

If you want to access data from the same project multiple times within a session, it may be useful to set the default project, rather than having to supply it every time. You can do this using `set_default_project()`. Then, you can just supply the endpoint, and the default project is used.

```{r}
set_default_project(mermaidr_project)
get_mermaid_project_endpoint("beltfishes")
```
