---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE
)
```
# mermaidr

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/data-mermaid/mermaidr.svg?branch=master)](https://travis-ci.com/data-mermaid/mermaidr)
<!-- badges: end -->

The goal of `mermaidr` is to access [MERMAID Collect](https://collect.datamermaid.org/) data directly from R.

## Installation

You can install mermaidr from GitHub with:

``` r
# install.packages("remotes")
remotes::install_github("data-mermaid/mermaidr@package", upgrade = "never")
```

Next, load the package:

```{r}
library(mermaidr)
```

## Authentication

`mermaidr` will help you interact with MERMAID Collect as an authenticated user, as soon as you need. This is only required for accessing project specific data. To access a list of all projects, sites, etc, you do not need to be authenticated.

If you would like to authenticate yourself immediately, use `mermaid_auth()`. This will open your browser to the MERMAID Collect login. Once you log in, you can go back to R and will be authenticated.

The login credentials expire every 24 hours. Once your credentials are expired, `mermaidr` will again help you automatically authenticate when needed.

## Usage

All functions in `mermaidr` are of the form `mermaid_*()`, to make functions easier to find and use when loaded with other packages!

## Accessing project data

To access data related to your MERMAID projects, first obtain a list of your projects with `mermaid_get_my_projects()`. 

At this point, you will have to authenticate to the Collect app. R will help you do this automatically by opening a browser window for you to log in to Collect, either via Google sign-in or username and password - however you normally do!

Once you've logged in, come back to R. Your login credentials will be stored for a day, until they expire, and you will need to login again. The package handles the expiration for you, so just log in again when prompted.

```{r}
library(mermaidr)
my_projects <- mermaid_get_my_projects()

my_projects
```

This function returns information on your projects, including project countries, the number of sites, tags, data policies, and more.

To filter for specific projects, you can use the `filter` function from `dplyr`:

```{r, message = FALSE}
library(dplyr)

indonesia_projects <- my_projects %>%
  filter(countries == "Indonesia")

indonesia_projects
```

Then, you can start to access data about your projects, like project sites via `mermaid_get_project_sites()`:

```{r}
indonesia_projects %>%
  mermaid_get_project_sites()
```

Or the managements for your projects via `mermaid_get_project_managements()`:

```{r}
indonesia_projects %>%
  mermaid_get_project_managements()
```

### Method data

You can also currently access data on your projects' fish belt and benthic PIT methods, with more methods to come. The details are in the following sections.

#### Fish Belt data

To access fish belt data for a project, use `mermaid_get_project_data()` with `method = "fishbelt"`. 

You can access individual observations (i.e., a record of each observation) by setting `data = "observations`":

```{r}
xpdc <- my_projects %>%
  filter(name == "XPDC Kei Kecil 2018")

xpdc %>%
  mermaid_get_project_data(method = "fishbelt", data = "observations")
```

You can access sample units data, which are observations aggregated to the sample units level. Fish belt sample units contain total biomass in kg/ha per sample unit, by trophic group:

```{r}
xpdc %>%
  mermaid_get_project_data("fishbelt", "sampleunits")
```

And finally, sample events data, which are aggregated further, to the sample event level. Fish belt sample events contain *mean* total biomass in kg/ha per sample event and by trophic group:

```{r}
xpdc %>%
  mermaid_get_project_data("fishbelt", "sampleevents")
```

#### Benthic PIT data

Accessing Benthic PIT data works the exact same way as Fish Belt data, except you change `method` to "benthicpit":

```{r}
xpdc %>%
  mermaid_get_project_data(method = "benthicpit", data = "observations")
```

You can access sample units and sample events the same way. 

For Benthic PIT, sample units contain percent cover per sample unit, by benthic category. Sample *events* contain *mean* percent cover per sample event, by benthic category. 

You can return both of these by setting the `data` argument. This will return a list of two data frames: one containing sample units, and the other sample events.

```{r}
xpdc_sample_units_events <- xpdc %>%
  mermaid_get_project_data(method = "benthicpit", data = c("sampleunits", "sampleevents"))

names(xpdc_sample_units_events)
xpdc_sample_units_events[["sampleunits"]]
```

#### Multiple methods data

To pull data for both fish belt and benthic PIT methods, you can set `method` to include both.

```{r}
xpdc_sample_events <- xpdc %>%
  mermaid_get_project_data(method = c("fishbelt", "benthicpit"), data = "sampleevents")
```

The result is a list of data frames, containing sample events for both fish belt and benthic PIT methods:

```{r}
names(xpdc_sample_events)

xpdc_sample_events[["benthicpit"]]
```

Alternatively, you can set `method` to "all" to pull for all methods! This will be handy once there are more than two methods. Similarly, you can set `data` to "all" to pull all types of data:

```{r}
all_project_data <- xpdc %>%
  mermaid_get_project_data(method = "all", data = "all", limit = 1)

names(all_project_data)

names(all_project_data[["benthicpit"]])
```


#### Multiple projects

Pulling data for multiple projects is the exact same, except there will be an additional "project" column at the beginning to distinguish which projects the data comes from. Recall that `my_projects contains four projects:

```{r}
my_projects
```

```{r}
my_projects %>%
  mermaid_get_project_data("fishbelt", "sampleevents", limit = 1)
```

Note the `limit` argument here, which just limits the data pulled to one record (per project, method, and data combination). This is useful if you want to get a preview of what your data will look like without having to pull it all in.

### Accessing non-project data

You may also want to access data that is not related to projects. To access this data, you do not need to authenticate R with MERMAID.

For example, you can also pull the reference data (the names and information of the fish and benthic attributes you can choose in MERMAID), using `mermaid_get_reference()`:

```{r}
mermaid_get_reference(reference = "fishfamilies")
```

Using this function, you can access the fish family, fish genera, fish species, and benthic attributes references by changing the `reference` argument.

You can also get a list of *all* projects (not just your own):

```{r}
mermaid_get_projects()
```

As well as all sites:

```{r}
mermaid_get_sites()
```

And all managements:

```{r}
mermaid_get_managements()
```

### Other data

There is additional data available from the MERMAID API, both related to specific projects and not. If you think you'll need to use these, please see the help for them by typing `?mermaid_get_endpoint` or `?mermaid_get_project_endpoint`.
