---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE
)
```
# mermaidr

<!-- badges: start -->
<!-- badges: end -->

The goal of `mermaidr` is to access [MERMAID Collect](https://collect.datamermaid.org/) data directly from R. The package is presently set to access the [dev MERMAID Collect](https://dev-collect.datamermaid.org/) only.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mermaidr/mermaidr@package")
```

## Authentication

`mermaidr` will help you interact with MERMAID Collect as an authenticated user, as soon as you need. This is only required for accessing project specific data. To access a list of projects, sites, etc, you do not need to be authenticated.

If you would like to authenticate yourself immediately, use `mermaid_auth()`. This will open your browser to the MERMAID Collect login. Once you log in, you can go back to R and will be authenticated. You will not need to do this again, so long as you tell R that you'd like to cache your token the first time around.

## Usage

To access the unauthenticated API endpoints, use `get_mermaid_endpoints()`. The results will return as a `tibble.` The following endpoints are available: "benthicattributes", "fishattributes", "fishfamilies", "fishgenera", "fishspecies", "managements", "projects", "sites".

For example,

```{r}
library(mermaidr)

get_mermaid_endpoint("projects")
```

```{r}
get_mermaid_endpoint("managements")
```

### Accessing project data

To access data for a specific project, use `get_mermaid_project_endpoint()` with the `project_id`, from `get_mermaid_endpoint("projects")`. This requires authentication, i.e. you will only be able to pull data for projects that you have access to.

The following project endpoints are available: "beltfishtransectmethods", "beltfishes", "benthiclittransectmethods", "benthicpittransectmethods", "benthicpits", "collectrecords", "habitatcomplexities", "obsbenthiclits", "obsbenthicpits", "obshabitatcomplexities", "obstransectbeltfishs", "managements", "observers", "project_profiles", "sampleevents", "sites".

For example:

```{r}
library(dplyr)

mermaidr_project <- search_projects("mermaidr testing", fixed = TRUE)

mermaidr_project %>%
  get_mermaid_project_endpoint("obshabitatcomplexities")
```

