---
title: "Adding Covariates to MERMAID data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accessing Project Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

TODO: this must be run on prod, due to sites used

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)

options(tibble.max_extra_cols = 20)
```

Prior to version 2.0.0, there were covariates on mermaidr data

Now if you try to access them, you get the message:

(run code with covariates = TRUE to see)

and the results are returned without covariates.

Need to use the `mermaidrcovariates` package.

Link to prescient, link to mermaidrcovariates package
https://mermaid.prescient.earth/


Previously included covariates:

# geomorphic zonation and benthic habitat from the \href{https://allencoralatlas.org}{Allen Coral Atlas}
# market gravity, water pollution (sediments and nitrogen), coastal population, industrial development (number of ports), tourism (reef value), and a cumulative pressure index from \href{https://conbio.onlinelibrary.wiley.com/doi/10.1111/conl.12858}{\emph{A global map of human pressures on tropical coral reefs}} by Andrello et al., 2022; 
# scores from \href{https://conbio.onlinelibrary.wiley.com/doi/10.1111/conl.12587}{\emph{Risk-sensitive planning for conserving coral reefs under rapid climate change}} by Beyer et al., 2018.

Still available:
# aca extent benthic habitat from allen coral atlas
# from andrello: market gravity, coastal population, number of ports, tourism, 
# from beyer: none

```{r}
library(mermaidr)
library(mermaidrcovariates)
library(tidyverse)

sites <- mermaid_get_sites(limit = 2000) %>%
  # TODO, update to project name once this is fixed
  filter(project == "d0ad9903-ad9d-4add-8ec3-22abeb8bb435")
  
sites %>%
  distinct(name)
```

```{r}
selected_covariates <- c("aca_benthic", "aca_extent", "human_pop_5km", "market_gravity", "num_ports_5km", "tourism_index")

# check which are raster and which are vector
purrr::map(selected_covariates, get_covariate_type)

# aca_benthic and aca_extent are raster, the rest are vector
raster_covariates <- c("aca_benthic", "aca_extent")
vector_covariates <- c("human_pop_5km", "market_gravity", "num_ports_5km", "tourism_index")
```

```{r}
# get raster covariates
raster_covariates_res_raw <- map(
  raster_covariates,
  \(covariate) get_zonal_statistics(sites, covariate, spatial_stats = "mean", radius = 100)
)

names(raster_covariates_res_raw) <- raster_covariates

raster_covariates_res_raw$aca_benthic %>%
  select(id, name, zonal_statistics)

raster_covariates_res_raw$aca_benthic %>%
  select(name, zonal_statistics) %>%
  unnest(zonal_statistics)

# do this in the map, just keep the value, and name it according to the index (covariate)

raster_covariates_res <- imap(
  raster_covariates_res_raw,
  \(res, covariate_name) {
    sites_with_covariate <- res %>%
      unnest(zonal_statistics) %>%
      select(names(sites), value)

    sites_with_covariate %>%
      pivot_longer(
        cols = value,
        names_to = "covariate",
        values_to = "value"
      ) %>%
      mutate(covariate = covariate_name)
  }
) %>%
  bind_rows()

# to make things easier, just keep ID and covariate/value, then join back on to sites

raster_covariates_res <- raster_covariates_res %>%
  select(name, covariate, value) %>%
  pivot_wider(names_from = covariate, values_from = value)

raster_covariates_res
```

```{r}
# do the same for vectors

# get vector covariates
vector_covariates_res_raw <- map(
  vector_covariates,
  \(covariate) attach_covariate_data(sites, covariate)
)

names(vector_covariates_res_raw) <- vector_covariates

vector_covariates_res_raw$human_pop_5km %>%
  select(name, pop_count, pop_count_raw)

vector_covariates_res <- imap(
  vector_covariates_res_raw,
  \(res, covariate_name) {
    covariate_cols <- res %>%
      select(-names(sites)) %>%
      names()

    res %>%
      pivot_longer(
        cols = all_of(covariate_cols),
        names_to = "covariate",
        values_to = "value"
      )
  }
) %>%
  bind_rows()

vector_covariates_res %>%
  select(name, covariate, value)

vector_covariates_res <- vector_covariates_res %>%
  select(name, covariate, value) %>%
  pivot_wider(names_from = covariate, values_from = value)

vector_covariates_res
```

Look at the two sets of covariates
```{r}
raster_covariates_res
vector_covariates_res
```

Join them back onto sites
```{r}
sites %>%
  left_join(raster_covariates_res, by = "name") %>%
  left_join(vector_covariates_res, by = "name")
```

